The effects of symbols and information in government communication on citizens  trust in policy: Evidence from a survey experiment

### 2018-05-19 paper.Rmd
Add files via upload

A growing body of public administration research seeks to assess whether and how government organizations can increse citizens' trust through direct communications with citizens. This study contributes to this discussion, by pointing to the influence of symbolic elements in public communications, and unpacking the theoretical mechanisms underlying their effect on citizens' trust. Using a rigorous experimental design, embedded in a realistic setting of communication of policy plans of the Israeli Environmental Protection Ministry, I tested the effect of symbolic elements on trust, and its interactions with the substantive content of the communications and with their perceived personal relevance. The experimental results demonstrate that using familiar symbolic elements in communications about policy plans increase citizens' trust in these policies, via two mechanisms: First, they evoke positive fellings that are transferred to the policy plan. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thus - they cause them to disregard its weaknesses, if exist. the latter mechanism entails an interaction between symbols and weak information. These theoretical mechanisms can be understood in light of the social psychology elaboration likelihood model. In addition, I theorized that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives, but the reults do not support this expectation. Rather, I found that citizens who perceived the policy issue as highly relevant for their lives (i.e. air-pollution in their area of residence), were similarly distracted by the graphic symbols, and persuaded to trust governmnet's actions, even when they were undoubtedly illogical. These findings are based on a large sample of more than `r (SVIVA2_01%>%nrow() %>% round(-2))-100` participants, and are robust for various model specifications.

What are the implications of these findings for public administation resarch? First, they highlight the important role of strategically designed symbolic elements in communications in shaping citizens' attitudes about government. Citizens' responses to symbolic elements and branding thus far received relatively little attention in Public administration literature, compared with the the abounded research on responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, this study demostrates that symbols decrease the "rational", systematic processing of substantive information. This may suggest that government organizations may bypass the cognitive biases that often undermine the effectiveness of positive information, by employing symbolic communication.[^footnote_biases_ELM] This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance. On the other hand, it challenges the democratic ideal that citizens' form their attitudes about governments on the basis of rational thinking [###Dahl...]. Moreover, it suggests that public organizations may use symbols in communications to compensate for their poor performance and inefficient policy plans. In turn, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which attenuates politicians' incentives to monitor and oversight them [##Busuioc_2016; Bertelli_2018##]. Overall... something about democratic accountability...

[^footnote_biases_ELM]: This posibility is supported, theoretically, by ELM. According to ELM, biased processing occures in the central route, whereby people give unproportional weight for strong or weak arguments. Accordingly, decresaing the ammount of thinking may also decrease biased processing [@petty_1986]. Future research might put this expectation to test.

Methodologically, these findings highlight that it is important to test the effectiveness of public communications, and not only the information.

Almost any government communication involves design elements, and many of them include familiar symbols and brand elements. Conversely, some of the current studies that explored the effect of information, did so by comparing participants' responses to pieces of information (mainly favorable vs. unfavorable performance) which were presented in surveys in the form of plain text and data tables without design elements which would be normally included in  such communications [@baekgaard_2016; @hvidman_2016; ]. This questions our ability to generalize from these experiments to citizens' responses to public communications in the real world.

Finally, PA can also implement ELM...

How can we explain the null findings about personal relevance %>%

limitations...

>>>>

A growing body of Public-Administration research seeks to assess whether and how government organizations can increase citizens' trust through direct communications with citizens. This study contributes to this discussion, by pointing to the influence of symbolic elements in public communications, and unpacking the theoretical mechanisms underlying their effect on citizens' trust. Using a rigorous experimental design, embedded in a realistic setting of the Israeli Environmental Protection Ministry's communication of policy plans, I tested the effect of symbolic elements on trust, and its interactions with the substantive content of the communications and with their perceived personal relevance. To the best of my knowledge, this is the first study to test these interactions in the context of government communications.

The experimental results demonstrate that using familiar symbolic elements in communications about policy plans increase citizens' trust in these policies, via two supplementary mechanisms: First, they evoke general positive feelings that are transferred to the specific policy plan. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thereby, they cause them to disregard its weaknesses. The second mechanism entails an interaction between symbols and weak information. These two mechanisms can be understood in light of the social psychology elaboration likelihood model. I theorized, based on ELM, that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives, but this expectation is not supported by the results. Rather, I found that citizens who perceived the policy issue as highly relevant for their lives (i.e. air-pollution in their area of residence), were similarly distracted by the symbols (and other peripheral cues), and persuaded to trust government's actions, even when they were undoubtedly illogical. These findings are based on a large sample of more than `r (SVIVA2_01%>%nrow() %>% round(-2))-100` participants, and are robust for various model specifications.

How can we explain the null findings about personal relevance?  
- very low motivation, even among Haifa-Bay subjects. Not new, existing problem. congruent with IEC study.

Limitation of study:
- one case...
- personal relevance based on observational.

What are the implications of these findings for public administration research? First, they highlight the important role of symbols and brands in shaping citizens' attitudes about government. Citizens' responses to symbolic elements in communications thus far received relatively little attention in Public administration literature, compared with the the abounded research on responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, this study shows that symbols decrease the "rational", systematic processing of substantive information. This may suggest that government organizations bypass or at least attenuate citizens' negative biases that undermine the effectiveness of positive information, by employing symbolic communication.[^footnote_biases_ELM] This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance. On the other hand, even if we assume that public organizations provide excellent outcomes, the emotive effect of government symbols on public opinion is still problematic, insofar as it challenges the democratic ideal that citizens form their opinions about their government and its actions on the basis of rational thinking [@dahl_1973; @soroka_2010]. Moreover, it becomes especially problematic when symbols are used by poorly performing organizations,  or organizations that lack professional expertiese. It those cases, symbolic elements can mitigate justified public criticism of government failures. Organizations can use them deliberately to distort public opinion via propaganda, or - more likely - without being aware of the tention between the symbols and their substantive message. Ultimately, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which provides them political power and attenuates elected politicians' incentives to monitor and oversight them [@busuioc_2016; @bertelli_2018]. Hence, government communications with citizens can be a tool for enhancing democratic responsiveness and accountability, yet they can also use it as a means to undermine them.

[^footnote_biases_ELM]: This possibility is supported, theoretically, by ELM. According to ELM, biased processing occurs in the central route, whereby people give proportional weight for strong or weak arguments. Accordingly, decreasing the amount of thinking may also decrease biased processing [@petty_1986]. Future research might put this expectation to test.

This study also contributes to existing behavioral Public Administration resarch by intoroducing ELM and applying it as a comprehensive theoretical framework for studying citizens' attitudes formation, and citizens' responses to government communications. ELM has various postulations and implications, but most importantly, it stresses that people respond differently to different communications, depending on people's motivation and ability to think about the subject, and on the quality of arguments and the peripheral cues in the communication. In this study, I applyed ELM to distinguish bewteen the peripheral processing of symbolic elements versus the central processing of substantive information, and to exlore their moderation by citizens' motivation to elaborate on the message. Future studies may apply ELM and further explore the effects of other peripheral cues in government communications, and of other relevant motivation and ability factors, to enhance our understanding of the conditions under which government communications are likely to be more/less effective.

In addition, ELM has important methodological implications regarding the external validity of survey experiment studies of citizens' responses to government communications. The participants of these studies, including this particular one, have relatively high motivation to elaborate on the information that scholars presented them, because they may assume that they will be asked about it afterwards. In case they receive payment or other compensation for their participation, this further increases their motivation. By comparison, in the "real world", citizens are far less motivated to invest in thinking about government messages, and they probably also have lower ability to do so, given the high competition on their time and attention. Accordingly, their responses to performance information will less affected by asymetrical processing of information, because most of them are not likely to process this information in the first place. Instead, most of them will rely on simple cues and heuristic shortcuts. One can postulate that these differences exacerbate biases due to partisan and political ideology [@james_2017a; @baekgaard_2016] and implicit attitudes about the public-sector [@marvel_2015; @marvel_2015a; @hvidman_2016], while attenuating the role of negativity bias [@nielsen_2016; @olsen_2017]. Future studies may empirically explore these interactions. Regardless, scholars should also take to account this methodological issue when designing survey experiments, and attempt to simulate real world conditions, as much as possible. For instance, they can add filler sections in the survey before the real manipulation to attenuate participants' motivation and ability to elaborate. In addition, because the graphic design of communications also affects motivation and ability to elborate, as shown here, scholars should simulate the design of real communications.

Closing...
